\documentclass{article}

\usepackage{hyperref}

\newcommand{\vcalc}{\textsc{VisiCalc}}

\title{Humanizing Learning Technology}
%\title{Humanizing Learning Technology\\with Actionable Educational Intelligence}
\author{Tom Nurkkala, PhD\\\texttt{tom.nurkkala@gmail.com}}

\begin{document}
\maketitle

\begin{abstract}
  Just as modern business software
  distills raw data into actionable management information,
  next generation learning management software
  should provide actionable educational intelligence
  that supports continuous, measurable improvement
  in teacher productivity and student outcomes.
  Furthermore, next generation learning management systems
  should re-personalize the relationship between teacher and student
  into a modern apprenticeship model
  that encourages mastery learning
  without sacrificing scalability
  to larger populations of students.
\end{abstract}

% \section{Introduction}
% \label{sec:introduction}

% This document captures thoughts
% about how to leverage, update, and create learning technology
% to nurture human interaction
% between student and teacher
% to the benefit of both.


\section{Limitations of Current Systems}
\label{sec:limitations}

Current and emerging learning management systems
represent a considerable improvement over the paper grade book,
but suffer two key limitations.
First,
they are \emph{not actionable}:
they provide little actionable information
that can be employed directly to engage students,
enhance learning outcomes,
and simplify course management.
Second,
they are \emph{not personalized}:
they emphasize massive scalability
at the expense of personalized interaction
between student and teacher.

\subsection{Not Actionable}
\label{sec:not-actionable}

Prior to the 1979 introduction
of \href{https://en.wikipedia.org/wiki/VisiCalc}{\vcalc{}},
the first computerized spreadsheet program,
business people used pencils and paper spreadsheets,
pre-printed with the now-universally-familiar rows and columns of cells.
Although \vcalc{} and its offspring
represented a huge advance over paper,
even modern spreadsheet software does little more
than automate the manual calculations
required by its paper progenitors.

By the 1990's, business software had gone beyond the spreadsheet,
embracing an important class of sophisticated systems
known collectively as Business Intelligence (BI) software.
While electronic spreadsheets
remain indispensable tools for organizing raw data,
BI systems transform data into actionable information
that can be used by business people
to analyze, manage, and empower their organization.

In educational technology,
the computerized learning management system (LMS)
represents an advance over the paper grade book
that is analogous to \vcalc{}'s superiority over the paper spreadsheet.
In addition to tracking scores and grades,
systems like
\href{https://www.blackboard.com/}{Blackboard},
\href{https://moodle.org/}{Moodle},
and
\href{https://www.instructure.com/canvas/}{Canvas}
are adept at distributing content,
tracking homework submissions,
mediating on-line ``conversations,''
and otherwise modernizing the learning environment.
But also like the electronic spreadsheet,
the current generation of LMS
does not go much beyond organizing raw data.
In particular,
the present-day LMS
does not provide either the student or the teacher
with significant actionable information
that helps analyze, manage, or empower the learning enterprise.

\subsection{Not Personalized}
\label{sec:not-personalized}

Current educational technologies
exhibit breathless enthusiasm
for scaling education to the very large.
Distance learning, on-line classes and degree programs,
and the ominously named Massive Open Online Course (MOOC)
make this tendency evident.
Examples abound from non-profit initiatives like
\href{https://www.edx.org/}{EdX} and
\href{https://www.khanacademy.org/}{Khan Academy}
and from for-profit companies like
\href{https://www.coursera.org/}{Coursera}
and
\href{https://www.udacity.com/}{Udacity},
not to mention the central place of technology
in the business models of for-profit institutions
like
\href{https://waldenu.edu/}{Walden} and the
\href{https://www.phoenix.edu/}{University of Phoenix}.

The emphasis on massively scalable educational opportunity
is understandable in an age of financial challenges in education,
large numbers of non-tradi\-tional students,
ubiquitous mobile technology,
global Internet access,
and billion-member online social networks.
But in the headlong rush
to deliver pre-packaged content to the masses,
we have neglected the historically significant relationship
between teacher and student
that has been the hallmark of deep learning since Socrates himself.
Significant, impactful, and long-lasting educational outcomes
are surely more likely to spring
from a face-to-face mentoring relationship
than from a YouTube video or an on-line chat room.

\section{Vision for a New System}
\label{sec:vision}

As the antidote to the
limitations enumerated in Section~\ref{sec:limitations},
the next generation of learning management system
should exhibit these interlocking characteristics:
\begin{enumerate}
\item \emph{Actionable}.
  It should distill raw learning data
  into educational intelligence
  on which students and (especially) teachers can act.
\item \emph{Personalized}.
  It should make explicit
  the learning needs of individual students,
  informing teachers in real time
  of the need for remediation
  or the opportunity for enhanced learning.
\item \emph{Scalable}.
  It should scale
  to a reasonably large numbers of students
  without unnecessarily burdening the teacher
  or compromising the foregoing characteristics.
\end{enumerate}

As a unifying motif
for the next generation of learning management system,
I propose a time-honored system of mastery learning: \emph{apprenticeship}.
In the medieval guild system,
a teenager was apprenticed to a master craftsman
to gain knowledge and develop skills
through direct exposure, extensive practice,
constant vigilance, and immediate feedback.
Apprenticeship was the essence of personalized education then,
and remains a viable model for effective education today.
This is not only true of vocational and technical schools.
Indeed, many undergraduate programs
require industry practica (a kind of ``apprenticeship light''),
and graduate education is itself
the clearest example of apprenticeship
in modern higher education.

Apprenticeship is highly personalized,
but its one-to-one ``faculty-student ratio''
appears to make it a hopelessly impractical anachronism.
Enter ``LMS~2.0'':
the next generation of learning management system
as a key enabling technology.
It can relieve teachers
of the mundane and time consuming details of course management.
It can provide actionable educational intelligence to teachers,
encouraging timely, strategic, and personal connection with students.
And it can scale to a large number of students.

Consider the following examples of capabilities such a system could deliver.

\paragraph{Course Management}
\begin{enumerate}
\item Rich tools for course planning and syllabus development, allowing
  topics and activities to be aligned explicitly with course learning outcomes
  and categorized according to Bloom's taxonomy.
\item Decoupling of course planning from course scheduling, allowing existing
  courses to be scheduled in future semesters with ease.
\item Automated production of course syllabi, based on course planning and
  scheduling information.
\item Topical overview prepared automatically for each class session, showing
  students the day's topics, preparatory readings, follow-up assignments,
  upcoming exams, etc. Daily topics would be tied explicitly to course learning
  outcomes to contextualize the day's learning.
\end{enumerate}

\paragraph{Classroom Experience}
\begin{enumerate}
\item Teacher and students make full use of mobile technology in the classroom,
  coordinated by interaction with a central server and database.
\item Automatic attendance taken when students ``log in to class.''
\item Speaker console decoupled from classroom display,
  encouraging meaningful classroom visuals,
  avoiding ``death by PowerPoint,''
  and discouraging the
  teacher's use of a slide deck as a teleprompter.
\item Ubiquitous use of mobile devices for classroom interaction; for example,
  in-class quizzing (``clickers on steroids''), mobile-based exercises by
  individual students or small student groups, real time monitoring by teacher
  of any student's work in progress, display of selected student work for
  classroom display and discussion.
\item Efficacy of teaching strategies measured empirically using in-class
  quizzing.
\item Student control of classroom graphics on personal mobile device, allowing
  review of material without disrupting class and consuming material at an
  individualized pace.
\item Server-based tracking of the pace of student engagement
  with course material throughout a class.
\item Student submission of private comments, requests, suggestions,
  or reminders to the teacher for attention either during or after class.
\end{enumerate}

\paragraph{Individualized Mentoring}
\begin{enumerate}
\item Invitations to office hours (or other remediation)
  for students who are lagging behind the class
  (e.g., as measured by whether they ``keep up''
  when viewing classroom graphics
  as they are covered by the teacher).
\item Remedial exercises for students
  struggling with in-class exercises or quizzes.
\item Automated relief from assigned homework for students demonstrating
  excellent performance on in-class exercises.
\item Referral to campus academic enrichment programs for students
  demonstrating possible learning challenges.
\end{enumerate}

\paragraph{Educational Analytics}
\begin{enumerate}
\item Course dashboard for teachers
  that distills raw data on attendance,
  in-class quiz results,
  students in need of personal attention, etc.
\item Graphical time line
  showing which students are ``keeping up''
  with in-class material,
  which students may need more attention,
  which are asleep during class, etc.
\item Comparative statistical analysis
  and graphical display
  of the efficacy of various learning activities
  in the classroom.
\item Analysis of time spent on each topic,
  tied to course planning information
  to help ensure class is on schedule
  and that schedules are realistic.
\item Information to help optimize the course
  for future semesters
  (e.g., content, pacing, student interaction, exercise relevance, etc.).
\end{enumerate}

None of these capabilities
require the development of radically new technologies.
The principle intellectual challenge
is to assemble \emph{existing} technologies
in ways that facilitate this actionable, personalized, and scalable vision
for teaching and learning.
The principle need that must be met to make this vision a reality
is adequate funding for a collaborative team with expertise in
teaching, curriculum development, computer science, software engineering,
user experience design, and information technology.

\section{Distance Learning and Pandemic}
\label{sec:pandemic}

The ideas outlined in Section~\ref{sec:vision}
were conceived with relation to
face-to-face classroom teaching and learning.
Due to the global pandemic,
many educational institutions
have turned to remote learning
or to hybrid models mixing in-person and remote instruction.

In addition to challenges posed by limited infrastructure
and resources along the digital divide,
a key problem with remote or hybrid learning
is the lack of interpersonal connection
between teachers and students.
The LMS~2.0 model proposed here
offers the promise of ameliorating
this problem.
By surfacing
actionable information on student performance,
teachers will be better able
to monitor and mentor students,
even when they are learning remotely.

The prevailing sentiment
among academic stakeholders
is a desire to return to the ``old normal.''
As the pandemic ebbs,
the academy will undoubtedly move
in this direction.
That said,
there will be stakeholders
who have found the
pivot to remote and hybrid instruction
to have advantages over the old normal.
Thus,
both for the foreseeable future
(shadowed by the pandemic)
and the longer term
(after the pandemic's eradication),
the tools and technologies
enumerated here
promise to be of
considerable and continuing
utility.

\section{Future Work}
\label{sec:future-work}

The features suggested in Section~\ref{sec:vision}
represent a small subset of would could be implemented
in a next generation learning management system.
As the system is developed and deployed in the classroom,
it can itself become the object of study,
allowing the LMS to be revised and refined
based on measurable empirical evidence
on student performance, learning outcomes, teacher experience, and so forth.

This empirical focus
should also provide ample opportunity for interdepartmental collaboration:
exploring the efficacy of the new learning management techniques
in diverse disciplines.
It will also be important to design studies
and evaluate results in collaboration
with the departments of Education, Psychology, and Sociology.

Sharing the results of the project with the scholarly community
will be possible in two ways.
First, publication of empirical results
derived from use of the system in Taylor classrooms.
Second, sharing the software itself
under a free and open-source software license
that allows other institutions
to deploy the system and
to participate in its development and expansion.

A long-term measure of the success and efficacy
of the project would be the emergent formation
of a global learning community
around continuous improvement of teaching and learning
through the judicious and sustainable application
of actionable, personalized learning management technology.

\section{Institutional Impact}
\label{sec:institutional-impact}

A next-generation learning management system would be a fertile source for
empirical research into educational practice across the disciplines, building
the University's reputation as a leader in innovative and effective pedagogy.

Although yet to be validated empirically, I expect measurable improvements in
student learning outcomes through the use of this technology. This result would
be of obvious benefit to students across the University.

\vfill
\begin{flushleft}
  \small
  Revision History
  \begin{itemize}
  \item 2017-10-01---Initial revision
  \item 2017-11-01---Sabbatical application submission
  \item 2020-10-06---Revised for general distribution
  \end{itemize}
\end{flushleft}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

% LocalWords:  pre EdX Coursera Udacity tradi tional
